---
title: "Practical Machine Learning Project"
author: "Orsolya Nyari"
date: "2026-04-30"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)
library(caret)
library(randomForest)
library(gbm)

knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

The goal of this project is to predict the manner in which participants performed barbell lifts using accelerometer data.

The outcome variable is **classe**, which represents five different execution styles (A–E).

# 2. Data Cleaning
We remove variables with too many missing values and irrelevant metadata.

```{r}
train <- read.csv("pml-training.csv")
test <- read.csv("pml-testing.csv")

train_clean <- train %>%
  select_if(~ sum(is.na(.)) < 0.5 * nrow(train)) %>%
  select(-c(X, user_name,
            raw_timestamp_part_1,
            raw_timestamp_part_2,
            cvtd_timestamp,
            new_window))

# Remove near-zero variance (VERY IMPORTANT SPEED BOOST)
nzv <- caret::nearZeroVar(train_clean)
train_clean <- train_clean[, -nzv]

```
# 3. Train/Test Split
We split the dataset into training and validation sets.

```{r}
set.seed(123)
inTrain <- caret::createDataPartition(train_clean$classe, p = 0.7, list = FALSE)

training <- train_clean[inTrain, ]
testing  <- train_clean[-inTrain, ]

```
# 4. Model1: Random Forest
Random Forest is used as a strong baseline model.
```{r}
set.seed(123)

rf_model <- train(
  classe ~ .,
  data = training,
  method = "rf",
  trControl = caret::trainControl(method = "cv", number = 3),
  ntree = 50
)

rf_pred <- predict(rf_model, testing)
rf_acc <- mean(rf_pred == testing$classe)

```
# 5. Model2: Gradient Boosting Machine (GBM)
GBM is used for comparison and tuning
```{r}
set.seed(123)

fitControl <- caret::trainControl(method = "cv", number = 3)

gbm_grid <- expand.grid(
  interaction.depth = c(3),
  n.trees = c(100),
  shrinkage = 0.1,
  n.minobsinnode = 10
)

gbm_model <- train(
  classe ~ .,
  data = training,
  method = "gbm",
  trControl = fitControl,
  tuneGrid = gbm_grid,
  verbose = FALSE
)

gbm_pred <- predict(gbm_model, testing)
gbm_acc <- mean(gbm_pred == testing$classe)

```
# 6. Model Comparison
```{r}
results <- data.frame(
  Model = c("Random Forest (fast)", "GBM (fast)"),
  Accuracy = c(rf_acc, gbm_acc),
  Error = c(1 - rf_acc, 1 - gbm_acc)
)
results

```
# 7. Variable Importance
The variable importance plot shows which features contribute most to the prediction. This helps interpret the model and understand which sensor measurements are most relevant.get
```{r}
plot(varImp(rf_model))
```

# 8. Final Model and Prediction
Random Forest is selected as the final model due to higher accuracy.
```{r echo=FALSE}
final_pred <- predict(rf_model, test)
final_pred

```
# 9. Conclusion
Both models performed well on this dataset due to strong signal in sensor data.

Random Forest achieved the highest accuracy and was selected as the final model.

The expected out-of-sample error is approximately 1%.
